Ownership types in theory and practice (in Rust)

### 2023-01-23 paper/paper.tex
edited domain specific ownership terms section in paper

In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used in this chapter, are defined.

The core concept of ownership types is that an object can be the \emph{owner} of other objects.

Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

In most ownership-type variants, the heap is divided into different regions called \emph{ownership contexts}.
Each object has one single ownership context, which is the set of all objects it owns.
Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.

>>>>

In the following, some terms from the Ownership Types survey \cite{ownership-types-survey}, which are commonly used in this chapter, are defined.

The core concept of ownership types is that each object is \emph{owned} by (at most) one other object, called its \emph{owner}.
Objects are organized into different \emph{ownership contexts}.
An ownership context is the set of all objects with the same owner.
Objects that are in the same ownership context, meaning they share the same owner, are called \emph{siblings} or \emph{peers}.
More informally, an object's ownership context can also be visualized as a box into which all of its owned objects are placed, to store and protect these objects.
Each object is considered to be \emph{inside} the object whose ownership context box it is placed in.
There also exists a more formal definition of the \emph{inside} relation:
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.
This hierarchy induces a nesting relationship between objects. \cite{ownership-types-survey}

### 2023-01-23 slides/slides.tex
edited domain specific ownership terms section in paper

\documentclass{beamer}

@@

\begin{frame}{Aliasing}
\begin{itemize}
  \item object aliasing: accessing the same memory through different symbolic names
  % TODO: example
  \item how to prevent bugs through (unintentional) aliases $\rightarrow$ bugs (hard to detect)
  \begin{itemize}
    \item ban aliasing altogether?
    \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Aliasing: example}
C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);    // v buffer reallocated => vptr dangling
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{frame}{Ownership types}
\begin{itemize}
  \item object-oriented programs:
  \begin{itemize}
    \item objects can reference any other object
    \item objects can read and modify other objects' fields
    \item makes it difficult to understand, to maintain, and to reason about
  \end{itemize}
  \item using ownership types:
  \begin{itemize}
    \item limit which objects can be referenced
    \item can specify whether the referenced objects may be mutated or just read from
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ownership types: core concepts}
\begin{itemize}
  % TODO: different highlighting than italics
  \item objects can be \emph{owners} of other objects
  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
  \item \emph{outside} relation: converse of \emph{inside} relation
  \item \emph{ownership context}: set of all objects that an object owns
  \item \emph{siblings}: objects in the same ownership context
  \item special \emph{world} object is root of the ownership hierarchy tree
\end{itemize}

@@

\begin{frame}{Owners-as-dominators: implications}
\begin{itemize}
  \item any external reference to an object must go through its owner
  \item objects are only protected from external access, not internal access
  \item owners-as-dominators: owner is dominator for all of its owned objects
  \item doesn't allow common idioms that involve aliasing
  \item model restricts \emph{where} references can point, not \emph{how} references are used
\end{itemize}

@@

\begin{frame}{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extends owners-as-modifiers
  \begin{itemize}
    \item no garbage collection (overhead)
    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
    \item rules for compiler to check validity of references
    \item extensions:
    \begin{itemize}
      \item transferring ownership from one object to another
      \item object immutability by default
      \item Resource Acquisition Is Initialization (similar to C++)
    \end{itemize}
  \end{itemize}
  % \item aliasing concept essential to ownership types
  % \item different ways of implementing aliasing:
  % \begin{itemize}
  %   \item pointers: exist in Rust, but their use is not advised
  %   \item \emph{references}
  % \end{itemize}
  \item ownership rules in Rust:
  \begin{enumerate}
    \item Each value in Rust has an owner.
    \item There can only be one owner at a time.
    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
  \end{enumerate}
\end{itemize}
\end{frame}

@@

  \begin{lstlisting}[language=Rust]
  let x: u32 = 42;
  let y: u32 = x; // x is copied instead of moved
  println!("x: {}, y: {}", x, y); // x and y valid
  \end{lstlisting}
  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Lifetimes}
\begin{itemize}
  \item every reference has a \emph{lifetime}
  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
  \item compiler constraints:
  \begin{enumerate}
    \item The reference can only be used while its lifetime is ongoing, and
    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Lifetimes: example}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  fn main() {
    let s1 = String::from("short");
    let s2 = String::from("longest");
    let longest = longest(&s1, &s2);
    println!("longest string: '{}'", longest);
  }
  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
    if s1.len() > s2.len() {
      s1
    } else {
      s2
    }
  }
  \end{lstlisting}
  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
  \item 's1' and 's2' have to live at least as long as lifetime 'a
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  println!("{} and {}", r1, r2);
  let r3 = &mut s;
  r3.push_str(" world");
  \end{lstlisting}

@@

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s; // lifetime 'a starts
  let r2 = &s; // lifetime 'b starts
  println!("{} and {}", r1, r2);
  // lifetimes 'a and 'b end
  let r3 = &mut s; // lifetime 'c starts
  r3.push_str(" world");
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

>>>>

\documentclass[handout]{beamer}

@@

% \begin{frame}{Aliasing}
% \begin{itemize}
%   \item 
%   % TODO: example
%   \item how to prevent bugs through unintentional aliases (hard to detect)?
%   \begin{itemize}
%     \item ban aliasing altogether
%     \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
%   \end{itemize}
% \end{itemize}
% \end{frame}

\begin{frame}[fragile]{Aliasing}
\emph{aliasing}: accessing same memory through different symbolic names
\pause

C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);    // v buffer reallocated => vptr dangling
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}
\pause

@@

\begin{frame}{Ownership types}
\begin{itemize}
  % \item object-oriented programs:
  % \begin{itemize}
  %   \item objects can reference any other object
  %   \item objects can read and modify other objects' fields
  %   \item makes it difficult to understand, to maintain, and to reason about
  % \end{itemize}
  \item using ownership types:
  \begin{itemize}
    \item limit which objects can be referenced
    \item specify whether the referenced objects may be mutated or just read from
    \item no garbage collection (overhead)
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ownership types: core concepts}
\begin{itemize}
  % TODO: different highlighting than italics
  \item objects can be \emph{owners} of other objects
  \item special \emph{world} object is root of the ownership hierarchy tree
  \item \emph{ownership context}: set of all objects that an object owns
  \item \emph{siblings}: objects in the same ownership context
  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
  \item \emph{outside} relation: converse of \emph{inside} relation
\end{itemize}

@@

\begin{frame}{Owners-as-dominators: implications}
\begin{itemize}
  \item any external reference to an object must go through its owner
  \item owners-as-dominators: owner is dominator for all of its owned objects
  \item objects are only protected from external access, not internal access
  \item model restricts \emph{where} references can point, not \emph{how} references are used
  \item doesn't allow common idioms that involve aliasing
\end{itemize}

@@

\begin{frame}{From theory to implementation}
\begin{itemize}
  \item Rust's ownership system: extends owners-as-modifiers
  \begin{itemize}
    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
    \item rules for compiler to check validity of references
    \item extensions:
    \begin{itemize}
      \item transferring ownership from one object to another
      \item object immutability by default
      % \item Resource Acquisition Is Initialization (similar to C++)
    \end{itemize}
  \end{itemize}
  % \item aliasing concept essential to ownership types
  % \item different ways of implementing aliasing:
  % \begin{itemize}
  %   \item pointers: exist in Rust, but their use is not advised
  %   \item \emph{references}
  % \end{itemize}
  \item ownership rules in Rust:
  \begin{enumerate}
    \item Each value in Rust has an owner.
    \item There can only be one owner at a time.
    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
  \end{enumerate}
\end{itemize}
\end{frame}

@@

  \begin{lstlisting}[language=Rust]
  let x: u32 = 42;
  let y: u32 = x; // x is copied instead of moved
  println!("x: {}, y: {}", x, y); // x and y valid
  \end{lstlisting}
  \item stack-based values implement \verb|Copy| trait
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Lifetimes}
\begin{itemize}
  \item every reference has a \emph{lifetime}: scope for which reference is valid
  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
  \item program is valid if all of the below hold for all references:
  \begin{enumerate}
    \item the reference can only be used while its lifetime is ongoing
    % \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
    \item immutable reference: the borrowed value does not get mutated during lifetime
    \item mutable reference: the borrowed value is not used at all during lifetime
  \end{enumerate}
  % \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  println!("{} and {}", *r1, *r2);
  let r3 = &mut s;
  r3.push_str(" world");
  \end{lstlisting}

@@

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s; // lifetime 'a starts
  let r2 = &s; // lifetime 'b starts
  println!("{} and {}", *r1, *r2);
  // lifetimes 'a and 'b end
  let r3 = &mut s; // lifetime 'c starts
  r3.push_str(" world");
  \end{lstlisting}
\end{itemize}
\end{frame}

@@

% \begin{frame}[fragile]{Lifetimes: example}
% \begin{itemize}
%   \begin{lstlisting}[language=Rust]
%   fn main() {
%     let s1 = String::from("short");
%     let s2 = String::from("longest");
%     let longest = longest(&s1, &s2);
%     println!("longest string: '{}'", longest);
%   }
%   fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
%     if s1.len() > s2.len() {
%       s1
%     } else {
%       s2
%     }
%   }
%   \end{lstlisting}
%   \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
%   \item 's1' and 's2' have to live at least as long as lifetime 'a
% \end{itemize}
% \end{frame}
